---
title: "Simulation analysis of brainscore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation analysis of brainscore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

This tutorial demonstrates how to perform do simulation to assess the false positive rate of analysis with brainscore. 
The `brainscore` function calculates the gene set scores and correlates them with the predictor variable. 
The simulation analysis is performed by 

(1) permuting the predictor variable (`sim_type ='randomize_pred'`);

(2) spinning the brain data (`sim_type ='spin_brain'`);

(3) resampling the gene data (`sim_type ='resampling_gene'`). 

The simulation analysis is performed to assess the false positive rate of the analysis, as when the above elements are permuted, we expect no the correlation between the gene set scores and the predictor variable.
For demonstration purposes, we have used a small number of iterations (sim_n = 5) and one subsample size (subsample_size = 100). In practice, we recommend using a larger number of iterations (i.e. 1000), various subsample sizes (e.g., 50, 100, 150, 200) and larger size of null models for a more robust analysis.

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(BrainEnrich)
library(dplyr)
library(enrichplot)
library(ggplot2)
```

## 1. Prepare Data for the Analysis
First, we will prepare the data needed for enrichment analysis. The data is simulated (not real data) from the HCP dataset, and we'll be using various pre-defined gene sets from SynGO and brain data for analysis.

### 1.1. Load sample data simulated from HCP data.
The data is simulated from HCP data. Let's load and inspect it.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
data(sample_df)
str(sample_df)
```
### 1.2. prepare brain_data for scoring
Next, we'll prepare the brain data by selecting the relevant columns and transforming the data for scoring.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
brain_data <- dplyr::select(sample_df, starts_with("L_")) %>% t()
colnames(brain_data) <- paste0("sub-", 1:ncol(brain_data))
str(brain_data)
```

### 1.3. Load gene expression data for the Analysis
We will now load gene expression data for the analysis. This data is based on a specific brain atlas (Desikan) and includes gene expression levels (correlation across donnors > 0.4) for the left hemisphere.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
gene_data <- get_geneExp(atlas = "desikan", rdonor = "r0.4", hem = "L")
str(gene_data)
```

### 1.4. Load SynGO for the Analysis
We will load SynGO terms that will be used in the enrichment analysis. This data represents gene sets associated with various molecular functions.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
annoData <- get_annoData(type = "SynGO")
geneSetList <- get_geneSetList(annoData)
length(geneSetList)
print(sprintf("Number of SynGO terms: %d", length(geneSetList)))
```

#### Filter the geneSetList (Optional)
Filtering the gene sets by size is optional and is shown here for demonstration purposes. This step is embedded in the brainenrich/brainscore function.
```{r echo=TRUE, eval=FALSE}
selected.gs <- filter_geneSetList(bg_genes = colnames(gene_data), geneSetList = geneSetList, minGSSize = 20, maxGSSize = 200)
print(sprintf("%d MF terms have gene size ranging between 20 and 200", length(selected.gs)))
```

## 2. Run the Simulation Analysis 
In this section, we will run the enrichment analysis using the prepared data. We permute the data in different ways and subsample it to examine the false positive rate.

### 2.1. Simulation analysis with randomized pred variable i.e., BMI (`sim_type ='randomize_pred'`)
Note: Since the gene set scores are already calculated, randomizing the predictor variable (BMI) offers limited information regarding the false positive rate of individual scoring. Basically, it is testing the false discovery rate of the correlation (i.e., lm model).
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "randomize_pred",
  cor_method = "pearson",
  aggre_method = "mean",
  minGSSize = 20,
  maxGSSize = 200
)
str(res)
```
The result is a list of lists, where each sublist contains the significance testing results for each term, separately for both with and without FDR correction. 1 denotes significant (p or fdr-p < 0.05), 0 denotes non-significant.
The results of fdr corrected p-values across gene sets (Dependent_vars) are also available in the result list. However, as we are expecting no significant resutls for each of gene set (i.e., Dependent_var), FDR correction across gene sets could be inappropriate.

### 2.2. Simulation analysis with spin_brain (`sim_type ='spin_brain'`)
We simulate the brain data by spinning it and then calculate the gene set scores before correlating them with the predictor variable. Both parametric and non-parametric tests (here, based on spin_brain) are used.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "spin_brain",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10,
  perm_id = perm_id_dk_lh_5000,
  minGSSize = 20,
  maxGSSize = 200
)
str(res)
```
The result is a list of lists, where each sublist contains the significance testing results for each term, separately for both with and without FDR correction. 1 denotes significant (p or fdr-p < 0.05), 0 denotes non-significant.
The results of fdr corrected p-values across gene sets (Dependent_vars) are also available in the result list. However, as we are expecting no significant resutls for each of gene set (i.e., Dependent_var), FDR correction across gene sets could be inappropriate.
Two types of statistical tests are performed: parametric (based on lm model itself; without "np") and non-parametric (based on the permutation of the brain data; with "np"). The permutation is done by spinning the brain data. 

### 2.3. Simulation analysis with spin_brain (`sim_type ='spin_brain'`) and pre-calculated null gsScore
As generation the null gsScore is time-consuming, we can use the pre-calculated null gsScore to speed up the simulation.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# pre-calculate the null gsScore
gsScore.spin_brain <- brainscore(
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  null_model = "spin_brain",
  n_perm = 10,
  perm_id = perm_id_dk_lh_5000,
  minGSSize = 20,
  maxGSSize = 200
)

cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  gsScoreList.null = gsScore.spin_brain,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "spin_brain",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10,
  perm_id = perm_id_dk_lh_5000,
  minGSSize = 20,
  maxGSSize = 200
)
str(res)
```
To ensure that the gsScore.spin_brain was generated with the same parameters and can be used in the function,
The attribute of gsScore.spin_brain should match the following argument of the brainscore.simulate: 
cor_method, aggre_method, n_perm, perm_id, minGSSize, maxGSSize.


### 2.4. Simulation analysis with resample_gene (`sim_type ='resample_gene'`)
We simulate the gene sets by resampling genes and then calculate the gene set scores before correlating them with the predictor variable. Both parametric and non-parametric tests (here, based on resample_gene) are used.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "resample_gene",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10,
  minGSSize = 20,
  maxGSSize = 200
)
str(res)
```
The result is a list of lists, where each sublist contains the significance testing results for each term, separately for both with and without FDR correction. 1 denotes significant (p or fdr-p < 0.05), 0 denotes non-significant.
The results of fdr corrected p-values across gene sets (Dependent_vars) are also available in the result list. However, as we are expecting no significant resutls for each of gene set (i.e., Dependent_var), FDR correction across gene sets could be inappropriate.
Two types of statistical tests are performed: parametric (based on lm model itself; without "np") and non-parametric (based on the permutation of gene sets; with "np"). The permutation is done by spinning the brain data. 


### 2.5. Simulation analysis with resample_gene (`sim_type ='resample_gene'`) and pre-calculated null gsScore
As generation the null gsScore is time-consuming, we can use the pre-calculated null gsScore to speed up the simulation.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# pre-calculate the null gsScore
gsScore.resample_gene <- brainscore(
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  null_model = "resample_gene",
  n_perm = 10,
  minGSSize = 20,
  maxGSSize = 200
)

cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  gsScoreList.null = gsScore.resample_gene,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "resample_gene",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10,
  minGSSize = 20,
  maxGSSize = 200
)
str(res)
```
Again, to ensure that the gsScore.resample_gene was generated with the same parameters and can be used in the function,
The attribute of gsScore.resample_gene should match the following argument of the brainscore.simulate: 
cor_method, aggre_method, n_perm, perm_id, minGSSize, maxGSSize.

