---
title: "Enrichment analysis of derived group-level statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brainenrich}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval=TRUE}
# if (!requireNamespace("BrainEnrich", quietly=TRUE)){
#   install.packages("BrainEnrich")
# }
# if (!requireNamespace("enrichplot", quietly=TRUE)){
#   BiocManager::install("enrichplot")
# }
library(BrainEnrich)
library(dplyr)
library(enrichplot)
library(ggplot2)
```

## 1. Prepare Data for the Analysis
In this section, we will prepare the data needed for enrichment analysis. The data is simulated from the HCP dataset, and we'll be using various pre-defined gene sets from GO-MF and brain data for analysis.

### 1.1. Load sample data frame. The data is simulated from HCP data.
The data is simulated from HCP data. Let's load and inspect it.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
data(sample_df)
head(sample_df)
colnames(sample_df)
```
### 1.2. prepare brain_data for scoring
Next, we'll prepare the brain data by selecting the relevant columns and transforming the data for scoring.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
brain_data=dplyr::select(sample_df, starts_with('L_'))%>%t()
colnames(brain_data)=paste0('sub-',1:ncol(brain_data))
head(brain_data)
```

### 1.3. Load gene expression data for the Analysis
We will now load gene expression data for the analysis. This data is based on a specific brain atlas (Desikan) and includes gene expression levels (correlation across donnors > 0.4) for the left hemisphere.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
gene_data=get_geneExp(atlas='desikan', rdonor='r0.4',hem='L')
str(gene_data)
```

### 1.4. Load SynGO for the Analysis
We will load SynGO terms that will be used in the enrichment analysis. This data represents gene sets associated with various molecular functions.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
annoData=get_annoData(type='SynGO')
geneSetList=get_geneSetList(annoData)
length(geneSetList)
print(sprintf('Number of SynGO terms: %d',length(geneSetList)))
```

#### Filter the geneSetList (not required for the analysis)
Filtering the gene sets by size is optional and is shown here for demonstration purposes. This step is embedded in the brainenrich/brainscore function.
```{r echo=TRUE, eval=FALSE}
selected.gs=filter_geneSetList(bg_genes=colnames(gene_data), geneSetList=geneSetList, minGSSize=20, maxGSSize=200)
print(sprintf('%d MF terms have gene size ranging between 20 and 200',length(selected.gs)))
```

## 2. Run the Simulation Analysis 
In this section, we will run the individual enrichment analysis using the prepared data.

### 2.1. get the row score of the brain_data (`set null_model='none'`)
