---
title: "Group-level Enrichment Analysis (brainenrich)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group-level Enrichment Analysis (brainenrich)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial demonstrates how to perform an enrichment analysis of derived group-level statistics using the `BrainEnrich` package. We will use effect size maps of Bipolar Disorders (BD) and gene expression data to identify significant gene sets.


## 0. Load the required libraries:

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(BrainEnrich)
library(dplyr)
library(ggplot2)
library(ggseg)
library(kableExtra)
```

## 1. Prepare Data for the Analysis

### 1.1. Load brain_data (effect size maps of Bipolar Disorders)

We will load brain data that contains the effect size maps of Bipolar Disorders. This data includes case-control comparisons of regional cortical thickness.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load brain data from the package
data(brain_data)
# Display the structure of the brain data
str(brain_data)
```

#### Show effect size map for BD

Visualize the effect size map to understand the regional differences in cortical thickness.

First, we define a plot function (this is not included in the package as it is not a core function).
```{r}
#' Plot Brain Data
#'
#' This function creates a brain plot using the ggplot2 and ggseg packages.
#'
#' df2plot A data frame containing the data to plot.
#' ats A character string indicating the atlas to use ('dx', 'dk', 'aseg').
#' what2plot A character string indicating the variable to plot ('statistic').
#' filterby A character string indicating the filter to apply ('p.value', 'p.adj', 'none').
#' title2show A character string indicating the title of the plot.
#' limit2show A numeric vector of length 2 indicating the limits for the color scale.
#' legend2show A character string indicating the legend title.
#' hide_legend A logical value indicating whether to hide the legend.
#' hem A character string indicating which hemisphere to plot ('both', 'left', 'right').
#' low A character string indicating the color for the low end of the scale.
#' mid A character string indicating the color for the midpoint of the scale.
#' high A character string indicating the color for the high end of the scale.
#' sufix2remove A character string indicating the suffix to remove from labels.
#' return A ggplot2 object.

plot_brain <- function(df2plot,
                       ats = c("dx", "dk", "aseg"),
                       what2plot = "statistic",
                       filterby = c("p.value", "p.adj", "none"),
                       title2show = "",
                       limit2show = c(-15, 15),
                       legend2show = "Stat",
                       hide_legend = FALSE,
                       hem = "both",
                       low = "steelblue1",
                       mid = "white",
                       high = "firebrick1",
                       sufix2remove = "_thickavg") {
  # Match arguments with allowed values
  ats <- match.arg(ats)
  filterby <- match.arg(filterby)

  # Apply filters based on p-values
  if (filterby == "p.value") {
    df2plot <- df2plot %>% filter(.data$p.value < 0.05)
  } else if (filterby == "p.adj") {
    df2plot <- df2plot %>% filter(.data$p.adj < 0.05)
  } else if (filterby == "none") {
    df2plot <- df2plot
  }

  # Modify labels and join data based on anatomical terms
  if (ats == "dx" || ats == "dk") {
    # ggseg exports atlas constructors as functions; call them to get atlas data.
    atlas <- "dk"
    atlas_data <- getExportedValue("ggseg", atlas)()
    df2plot <- df2plot %>%
      mutate(
        label = sub(".*L_", "lh_", .data$label),
        label = sub(".*R_", "rh_", .data$label),
        label = sub(sufix2remove, "", .data$label)
      ) %>%
      brain_join(atlas_data) %>%
      reposition_brain(as.formula(". ~ hemi + view"))
  } else if (ats == "aseg") {
    atlas_data <- getExportedValue("ggseg", "aseg")()
    df2plot <- df2plot %>%
      mutate(
        label = sub("SV_L_", "Left-", .data$label), # Replace "SV_L_" with "Left-"
        label = sub("SV_R_", "Right-", .data$label), # Replace "SV_R_" with "Right-"
        label = sub("thal", "Thalamus-Proper", .data$label), # Replace "thal" with "Thalamus-Proper"
        label = sub("caud", "Caudate", .data$label), # Replace "caud" with "Caudate"
        label = sub("put", "Putamen", .data$label), # Replace "put" with "Putamen"
        label = sub("pal", "Pallidum", .data$label), # Replace "pal" with "Pallidum"
        label = sub("hippo", "Hippocampus", .data$label), # Replace "hippo" with "Hippocampus"
        label = sub("amyg", "Amygdala", .data$label), # Replace "amyg" with "Amygdala"
        label = sub("accumb", "Accumben-area", .data$label), # Replace "accumb" with "Accumben-area"
        label = sub("LatVent", "Lateral-Ventricle", .data$label) # Replace "LatVent" with "Lateral-Ventricle"
      ) %>%
      filter(!grepl("Accumben-area", .data$label)) %>%
      brain_join(atlas_data) %>%
      filter(.data$view == "coronal")
  }

  # Apply conditional filtering based on hemisphere
  if (hem %in% c("left", "right")) {
    df2plot <- df2plot %>% filter(.data$hemi == hem)
  }

  # Construct the plot
  p <- df2plot %>%
    ggplot() +
    geom_sf(aes(fill = .data[[what2plot]])) +
    scale_fill_gradient2(midpoint = 0, low = low, mid = mid, high = high, space = "Lab", limits = limit2show) +
    ggtitle(title2show) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(fill = legend2show) +
    {
      if (hide_legend) theme(legend.position = "none")
    }

  return(p)
}
```

```{r fig.align='center', fig.height=6, fig.width=6, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
df2plot <- brain_data %>% tibble::rownames_to_column("label")
plot_brain(df2plot,
  ats = "dk", filterby = "none", limit2show = c(-0.3, 0.3),
  what2plot = "MDD", hem = "left", low = "#0197b2", mid = "white", high = "orange",
  legend2show = "Effect Size"
)
```

### 1.2. Create perm_id for the Analysis
perm_id is a list of permutations of brain regions with a spinning approach while trying to maintain the spatial structure of the brain.
It is used to generate null distributions for the enrichment analysis.
If you use this function to create perm_ids, consider citing the original source at https://github.com/frantisekvasa/rotate_parcellation and relevant papers.

```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# read the centroid coordinates of the brain regions
data(coord_dk_lh)
perm_id_dk_lh_5000 <- rotate_parcellation(coord.l = coord_dk_lh, nrot = 5000, seed = 2024)
```

#### Here, we load the precomputed perm_id directly from the package.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
data(perm_id_dk_lh_5000)
str(perm_id_dk_lh_5000)
```

### 1.3. Load gene expression data for the Analysis

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
gene_data <- get_geneExp(atlas = "desikan", rdonor = "r0.4", hem = "L")
str(gene_data)
```

### 1.4. Load GO MF for the Analysis

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
annoData <- get_annoData(type = "GO_MF")
geneSetList <- get_geneSetList(annoData)
length(geneSetList)
print(sprintf("Number of MF terms: %d", length(geneSetList)))
```

#### Filter the geneSetList (optional). 
#### This step is embedded in the `brainenrich` function. Here is just to show how it is filtered.

```{r echo=TRUE, eval=FALSE}
selected.gs <- filter_geneSetList(bg_genes = colnames(gene_data), geneSetList = geneSetList, minGSSize = 20, maxGSSize = 200)
print(sprintf("%d MF terms have gene size ranging between 20 and 200", length(selected.gs)))
```

## 2. Run the Analysis 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
res <- brainenrich(
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  perm_id = perm_id_dk_lh_5000,
  cor_method = "pearson",
  aggre_method = "mean",
  null_model = "spin_brain",
  n_perm = 10,
  n_cores = 1,
  minGSSize = 20,
  maxGSSize = 200,
  pvalueCutoff = 1,
  threshold_type = "sd",
  threshold_value = 1
)
```

## 3. Visualize the Results

`brainenrich()` returns an `EnrichRes` object. The plotting helpers below work directly on the native result object.

### 3.1. Dot plot of top terms

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align='center', fig.height=6, fig.width=10}
plot_terms(res, type = "dot", top_n = 15)
```

### 3.2. Null-score ridge view

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_terms(res, type = "ridge", top_n = 10)
```

### 3.3. Heatmap of null scores

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_terms(res, top_n = 10, max_perm = 10)
```

### 3.4. Lightweight term network

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align='center', fig.height=6, fig.width=12}
plot_term_network(res, top_n = 10, min_overlap = 0.05)
```

### 3.5. Core genes for the top term

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align='center', fig.height=6, fig.width=8}
top_term <- res$table$ID[[1]]
plot_core_genes(res, term_id = top_term, mode = "impact")
```

### 4. Put results in a Table
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
res.df=as.data.frame(res)
# make scollable table
kable(res.df, format = "html", escape = FALSE) %>%
  kable_styling("striped", full_width = FALSE) %>%
  scroll_box(width = "100%", height = "400px")
