---
title: "Simulation analysis of brainscore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brainenrich}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval=TRUE}
# if (!requireNamespace("BrainEnrich", quietly=TRUE)){
#   install.packages("BrainEnrich")
# }
# if (!requireNamespace("enrichplot", quietly=TRUE)){
#   BiocManager::install("enrichplot")
# }
library(BrainEnrich)
library(dplyr)
library(enrichplot)
library(ggplot2)
```

## 1. Prepare Data for the Analysis
In this section, we will prepare the data needed for enrichment analysis. The data is simulated from the HCP dataset, and we'll be using various pre-defined gene sets from SynGO and brain data for analysis.

### 1.1. Load sample data frame. The data is simulated from HCP data.
The data is simulated from HCP data. Let's load and inspect it.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
data(sample_df)
head(sample_df)
colnames(sample_df)
```
### 1.2. prepare brain_data for scoring
Next, we'll prepare the brain data by selecting the relevant columns and transforming the data for scoring.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
brain_data <- dplyr::select(sample_df, starts_with("L_")) %>% t()
colnames(brain_data) <- paste0("sub-", 1:ncol(brain_data))
head(brain_data)
```

### 1.3. Load gene expression data for the Analysis
We will now load gene expression data for the analysis. This data is based on a specific brain atlas (Desikan) and includes gene expression levels (correlation across donnors > 0.4) for the left hemisphere.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
gene_data <- get_geneExp(atlas = "desikan", rdonor = "r0.4", hem = "L")
str(gene_data)
```

### 1.4. Load SynGO for the Analysis
We will load SynGO terms that will be used in the enrichment analysis. This data represents gene sets associated with various molecular functions.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
annoData <- get_annoData(type = "SynGO")
geneSetList <- get_geneSetList(annoData)
length(geneSetList)
print(sprintf("Number of SynGO terms: %d", length(geneSetList)))
```

#### Filter the geneSetList (not required for the analysis)
Filtering the gene sets by size is optional and is shown here for demonstration purposes. This step is embedded in the brainenrich/brainscore function.
```{r echo=TRUE, eval=FALSE}
selected.gs <- filter_geneSetList(bg_genes = colnames(gene_data), geneSetList = geneSetList, minGSSize = 20, maxGSSize = 200)
print(sprintf("%d MF terms have gene size ranging between 20 and 200", length(selected.gs)))
```

## 2. Run the Simulation Analysis 
In this section, we will run the enrichment analysis using the prepared data. We permute the data in different ways and subsample it to examine the false positive rate.
### 2.1. Simulation analysis with randomized pred variable i.e., BMI (`sim_type ='randomize_pred'`)
Note: Since the gene set scores are already calculated, randomizing the predictor variable (BMI) offers limited information regarding the false positive rate of individual scoring.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "randomize_pred",
  cor_method = "pearson",
  aggre_method = "mean"
)
head(str(res))
```
The result is a list of lists, where each sublist contains the significance testing results for each term, separately for both with and without FDR correction. 1 denotes significant, 0 denotes non-significant.
For demonstration purposes, we have used a small number of iterations (sim_n = 5) and one subsample size (subsample_size = 100). In practice, we recommend using a larger number of iterations (i.e. 1000) and various subsample sizes (e.g., 50, 100, 150, 200) for a more robust analysis.

### 2.2. Simulation analysis with spin_brain (`sim_type ='spin_brain'`)
We simulate the brain data by spinning it and then calculate the gene set scores before correlating them with the predictor variable. Both parametric and non-parametric tests (based on spin_brain) are used.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "spin_brain",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10,
  perm_id = perm_id_dk_lh_5000
)
str(res)
```
The result is a list of lists, where each sublist contains the significance testing results for each iteration. There are results separately for (FDR or no-FDR, parametric or non-parametric). 1 denotes significant, 0 denotes non-significant.
For demonstration purposes, we have used a small number of iterations (sim_n = 5), one subsample size (subsample_size = 100) and small permutation numer (n_perm=10). In practice, we recommend using a larger number of iterations (i.e. 1000), various subsample sizes (e.g., 50, 100, 150, 200) and more permutation times (e.g., perm_n=5000)for a more robust analysis.


### 2.3. Simulation analysis with spin_brain (`sim_type ='resample_gene'`)
We simulate the gene sets by resampling genes and then calculate the gene set scores before correlating them with the predictor variable. Both parametric and non-parametric tests (based on resample_gene) are used.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cov_df <- sample_df %>% dplyr::select(Age, Sex)
pred_df <- sample_df %>% dplyr::select(BMI)
res <- brainscore.simulate(
  pred_df = pred_df,
  cov_df = cov_df,
  brain_data = brain_data,
  gene_data = gene_data,
  annoData = annoData,
  sim_n = 5,
  subsample_size = 100,
  sim_type = "resample_gene",
  cor_method = "pearson",
  aggre_method = "mean",
  n_perm = 10
)
str(res)
```
The result is a list of lists, where each sublist contains the significance testing results for each iteration. There are results separately for (FDR or no-FDR, parametric or non-parametric). 1 denotes significant, 0 denotes non-significant.
For demonstration purposes, we have used a small number of iterations (sim_n = 5), one subsample size (subsample_size = 100) and small permutation numer (n_perm=10). In practice, we recommend using a larger number of iterations (i.e. 1000), various subsample sizes (e.g., 50, 100, 150, 200) and more permutation times (e.g., perm_n=5000)for a more robust analysis.
