---
title: "How to prepare gene sets (annoDat) for the analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to prepare gene sets (annoDat) for the analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=T, eval=F}
# Load required libraries
library(readxl)
library(dplyr)
library(tidyr)
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

bioc_packages <- c("org.Hs.eg.db", "clusterProfiler", "DOSE", "ReactomePA", "reactome.db")
for (pkg in bioc_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}
library(org.Hs.eg.db)
library(clusterProfiler)
library(DOSE)
library(ReactomePA)
library(reactome.db)
library(BrainEnrich)
build_Anno <- getFromNamespace("build_Anno", "DOSE")
```

As we may need to convert geneID to symbols, we first define helper functions for doing so
```{r, include = FALSE}
symbol_annoData <- function(annoData) {
  require(org.Hs.eg.db)
  tmp <- split_Anno(annoData)
  path2gene <- tmp$path2gene
  path2name <- tmp$path2name
  geneIDs <- path2gene$geneID
  entrez_ids <- AnnotationDbi::keys(org.Hs.eg.db, keytype = "ENTREZID")
  matching_ids <- geneIDs %in% entrez_ids
  percentage_match <- sum(matching_ids) / length(geneIDs) * 100
  threshold <- 90 # For example, 80% match
  is_mostly_entrez <- percentage_match >= threshold
  if (is_mostly_entrez) {
    gene_symbols <- mapIds(org.Hs.eg.db, keys = geneIDs, column = "SYMBOL", keytype = "ENTREZID", multiVals = "first")
  } else {
    stop("The gene IDs are not mostly ENTREZID. Attempting to convert to ENTREZID...")
  }
  if (length(gene_symbols) != length(geneIDs)) {
    stop("Failed to convert gene IDs in the data frame to SYMBOL.")
  }
  path2gene$geneID <- gene_symbols
  path2gene <- path2gene[!(is.na(path2gene$geneID) | path2gene$geneID == ""), ]
  build_Anno <- getFromNamespace("build_Anno", "DOSE")
  annoData <- build_Anno(path2gene, path2name)
  return(annoData)
}


# #' Convert Entrez IDs to Gene Symbols
# #'
# #' This function converts a vector of Entrez IDs to gene symbols using the org.Hs.eg.db annotation package.
# #'
# #' @param entrezid A vector of Entrez IDs to be converted to gene symbols.
# #' @return A vector of gene symbols corresponding to the input Entrez IDs.
# #' @importFrom AnnotationDbi mapIds
# #' @export
# entrezid2symbol <- function(entrezid) {
#   # Ensure input is character vector
#   entrezid <- as.character(entrezid)

#   # Map Entrez IDs to gene symbols
#   mappedSymbol <- suppressMessages(
#     mapIds(
#       x = org.Hs.eg.db,
#       keys = entrezid,
#       keytype = "ENTREZID",
#       column = "SYMBOL",
#       multiVals = "first"
#     )
#   )
#   # Remove NA values and attributes
#   mappedSymbol <- na.omit(mappedSymbol)
#   attributes(mappedSymbol) <- NULL
#   return(unname(mappedSymbol))
# }
```

# Gene ontology (GO) gene sets
```{r include=T, eval=F}
get_GO_data <- getFromNamespace("get_GO_data", "clusterProfiler")
annoData <- get_GO_data("org.Hs.eg.db", ont = "BP", keytype = "SYMBOL")
# saveRDS(annoData, file = "extdata/geneSets/GO_BP.rds", compress = "xz")
annoData <- get_GO_data("org.Hs.eg.db", ont = "MF", keytype = "SYMBOL")
# saveRDS(annoData, file = "extdata/geneSets/GO_MF.rds", compress = "xz")
annoData <- get_GO_data("org.Hs.eg.db", ont = "CC", keytype = "SYMBOL")
saveRDS(annoData, file = "inst/extdata/geneSets/GO_CC.rds", compress = "xz")


# check version of org.Hs.eg.db
packageVersion("clusterProfiler")
packageVersion("org.Hs.eg.db")
```

# DisGeNET gene sets
```{r include=T, eval=F}
get_DGN_data <- getFromNamespace("get_DGN_data", "DOSE")
annoData <- get_DGN_data()
test1 <- get_geneSetList(annoData)
str(head(test1))
get_termDescription("C0001546", annoData)
annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))

saveRDS(annoData, file = "inst/extdata/geneSets/DGN.rds", compress = "xz")
packageVersion("DOSE")
packageVersion("AnnotationDbi")
```

# KEGG modules
```{r include=T, eval=F}
prepare_KEGG <- getFromNamespace("prepare_KEGG", "clusterProfiler")
annoData <- prepare_KEGG("hsa", "MKEGG", "ncbi-geneid")
test1 <- get_geneSetList(annoData)
str(head(test1))

annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))
get_termDescription("M00124", annoData)
saveRDS(annoData, file = "inst/extdata/geneSets/MKEGG.rds", compress = "xz")

packageVersion("clusterProfiler")
```


# KEGG pathways
```{r include=T, eval=F}
prepare_KEGG <- getFromNamespace("prepare_KEGG", "clusterProfiler")
annoData <- prepare_KEGG("hsa", "KEGG", "ncbi-geneid")
test1 <- get_geneSetList(annoData)
str(head(test1))

annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))
get_termDescription("hsa03008", annoData)
saveRDS(annoData, file = "inst/extdata/geneSets/KEGG.rds", compress = "xz")
packageVersion("clusterProfiler")
```


# WikiPathways
```{r include=T, eval=F}
prepare_WP_data <- getFromNamespace("prepare_WP_data", "clusterProfiler")
wpdata <- prepare_WP_data("Homo sapiens")
TERM2GENE <- wpdata$WPID2GENE
TERM2NAME <- wpdata$WPID2NAME
build_Anno <- getFromNamespace("build_Anno", "DOSE")
annoData <- build_Anno(TERM2GENE, TERM2NAME)

test1 <- get_geneSetList(annoData)
str(head(test1))

annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))
get_termDescription("WP2338", annoData)
saveRDS(annoData, file = "inst/extdata/geneSets/WikiPathways.rds", compress = "xz")
packageVersion("clusterProfiler")
```

# Reactome
```{r include=T, eval=F}
get_Reactome_DATA <- getFromNamespace("get_Reactome_DATA", "ReactomePA")
annoData <- get_Reactome_DATA("human")
test1 <- get_geneSetList(annoData)
str(head(test1))

annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))
get_termDescription("R-HSA-1059683", annoData)
saveRDS(annoData, file = "inst/extdata/geneSets/Reactome.rds", compress = "xz")
packageVersion("ReactomePA")
```

# SynGO (how to make annoDat from xlsx)
```{r include=T, eval=F}
# Define URL for downloading SynGO data
url <- "https://www.syngoportal.org/data/SynGO_bulk_download_release_20231201.zip"
# Create temporary file and directory
zip_path <- tempfile(fileext = ".zip")
temp_dir <- tempdir()
# Clean-up function to ensure temp files are removed
on.exit(
  {
    if (file.exists(zip_path)) unlink(zip_path)
    if (file.exists(extracted_file)) unlink(extracted_file)
  },
  add = TRUE
)
extracted_file <- file.path(temp_dir, "syngo_ontologies.xlsx")
download.file(url, zip_path, mode = "wb")
unzip(zip_path, files = "syngo_ontologies.xlsx", exdir = temp_dir)


# Read the data from the extracted file
data <- read_xlsx(extracted_file)

# Process TERM2GENE
TERM2GENE <- data %>%
  dplyr::select(id, hgnc_symbol) %>%
  dplyr::rename(term = id, gene = hgnc_symbol) %>%
  dplyr::mutate(gene = strsplit(gene, ", ")) %>%
  unnest(cols = c(gene))

# Process TERM2NAME
TERM2NAME <- data %>%
  dplyr::select(id, name) %>%
  dplyr::rename(term = id, description = name)

# Use build_Anno function from DOSE package
build_Anno <- getFromNamespace("build_Anno", "DOSE")
annoData <- build_Anno(TERM2GENE, TERM2NAME)
# saveRDS(annoData, file = "extdata/geneSets/SynGO.rds",compress='xz')
```


# CellTypes Seidlitz2020 (how to prepare annoData with a csv file)
### Seidlitz, J., Nadig, A., Liu, S., Bethlehem, R. A., Vértes, P. E., Morgan, S. E., ... & Raznahan, A. (2020). Transcriptomic and cellular decoding of regional brain vulnerability to neurogenetic disorders. Nature communications, 11(1), 3358.
```{r include=T, eval=F}
library(dplyr)
# Define URL for Seidlitz2020
url <- "https://github.com/jms290/PolySyn_MSNs/blob/master/Data/AHBA/celltypes_PSP.csv?raw=true"

# Create temporary file
temp_file <- tempfile()
# Clean-up function to ensure temp files are removed
on.exit(
  {
    if (file.exists(temp_file)) unlink(temp_file)
  },
  add = TRUE
)

download.file(url, temp_file, mode = "wb")

# Read CSV file
TERM2GENE <- read.csv(temp_file) %>%
  dplyr::mutate(term = class) %>%
  filter(gene != "") %>%
  dplyr::select(term, gene)

key_dict <- c(
  "Astro" = "Astrocytes",
  "Endo" = "Endothelial",
  "Micro" = "Microglial",
  "OPC" = "Oligodendrocyte Progenitors",
  "Neuro-Ex" = "Excitatory Neurons",
  "Neuro-In" = "Inhibitory Neurons",
  "Oligo" = "Oligodendrocytes"
)

TERM2NAME <- TERM2GENE %>%
  dplyr::mutate(description = dplyr::recode(term, !!!key_dict)) %>%
  dplyr::select(term, description)
annoData <- build_Anno(TERM2GENE, TERM2NAME)
test1 <- get_geneSetList(annoData)
str(head(test1))
get_termDescription(term = "Astro", annoData)

saveRDS(annoData, file = "inst/extdata/geneSets/CellTypes_Seidlitz2020.rds", compress = "xz")
```


# CellTypes Lake2018 (how to prepare annoData from a gmt file)
### Lake, B. B., Chen, S., Sos, B. C., Fan, J., Kaeser, G. E., Yung, Y. C., ... & Zhang, K. (2018). Integrative single-cell analysis of transcriptional and epigenetic states in the human adult brain. Nature biotechnology, 36(1), 70-80.
```{r include=T, eval=F}
# Define URL for Lake2018
url <- "https://github.com/molecular-neuroimaging/Imaging_Transcriptomics/raw/main/imaging_transcriptomics/data/geneset_LAKE.gmt"
# Create temporary file
temp_file <- tempfile()
# Clean-up function to ensure temp files are removed
on.exit(
  {
    if (file.exists(temp_file)) unlink(temp_file)
  },
  add = TRUE
)


download.file(url, temp_file, mode = "wb")
# Read GMT file
TERM2GENE <- suppressWarnings({
  clusterProfiler::read.gmt(temp_file) %>%
    filter(gene != "") %>%
    dplyr::select(term, gene)
})

key_dict <- c(
  # Excitatory Neurons
  "Ex1" = "Excitatory Neurons 1",
  "Ex2" = "Excitatory Neurons 2",
  "Ex3a" = "Excitatory Neurons 3a",
  "Ex3b" = "Excitatory Neurons 3b",
  "Ex3c" = "Excitatory Neurons 3c",
  "Ex3d" = "Excitatory Neurons 3d",
  "Ex3e" = "Excitatory Neurons 3e",
  "Ex4" = "Excitatory Neurons 4",
  "Ex5a" = "Excitatory Neurons 5a",
  "Ex5b" = "Excitatory Neurons 5b",
  "Ex6a" = "Excitatory Neurons 6a",
  "Ex6b" = "Excitatory Neurons 6b",
  "Ex8" = "Excitatory Neurons 8",

  # Granule Cells (Cerebellar)
  "Gran" = "Cerebellar Granule Cells",

  # Inhibitory Neurons
  "In1a" = "Inhibitory Neurons 1a",
  "In1b" = "Inhibitory Neurons 1b",
  "In1c" = "Inhibitory Neurons 1c",
  "In2" = "Inhibitory Neurons 2",
  "In3" = "Inhibitory Neurons 3",
  "In4a" = "Inhibitory Neurons 4a",
  "In4b" = "Inhibitory Neurons 4b",
  "In6a" = "Inhibitory Neurons 6a",
  "In6b" = "Inhibitory Neurons 6b",
  "In7" = "Inhibitory Neurons 7",
  "In8" = "Inhibitory Neurons 8",

  # Purkinje Cells (Cerebellar Neurons)
  "Purk1" = "Purkinje Cells 1",
  "Purk2" = "Purkinje Cells 2",

  # Non-Neuronal Cells
  "End" = "Endothelial Cells",
  "Per" = "Pericytes/Smooth Muscle Cells",
  "Ast" = "Astrocytes",
  "Ast_Cer" = "Cerebellar Astrocytes",
  "Oli" = "Oligodendrocytes",
  "OPC" = "Oligodendrocyte Progenitor Cells",
  "OPC_Cer" = "Cerebellar OPCs",
  "Mic" = "Microglia"
)




TERM2NAME <- TERM2GENE %>%
  dplyr::mutate(description = dplyr::recode(term, !!!key_dict)) %>%
  dplyr::select(term, description)
unique(TERM2NAME)

annoData <- build_Anno(TERM2GENE, TERM2NAME)
test1 <- get_geneSetList(annoData)
str(head(test1))
get_termDescription(term = "Ast", annoData)

saveRDS(annoData, file = "inst/extdata/geneSets/CellTypes_Lake2018.rds", compress = "xz")
```

# CellTypes Martins2021 (how to prepare annoData with a gmt file)
### Imaging transcriptomics: Convergent cellular, transcriptomic, and molecular neuroimaging signatures in the healthy adult human brain. Daniel Martins, Alessio Giacomel, Steven CR Williams, Federico Turkheimer, Ottavia Dipasquale, Mattia Veronese, PET templates working group. Cell Reports.
```{r include=T, eval=F}
# Define URL for Martins2021
url <- "https://github.com/molecular-neuroimaging/Imaging_Transcriptomics/raw/main/imaging_transcriptomics/data/geneset_Pooled.gmt"
# Create temporary file
temp_file <- tempfile()
# Clean-up function to ensure temp files are removed
on.exit(
  {
    if (file.exists(temp_file)) unlink(temp_file)
  },
  add = TRUE
)
download.file(url, temp_file, mode = "wb")
# Read GMT file
TERM2GENE <- suppressWarnings({
  clusterProfiler::read.gmt(temp_file) %>%
    filter(gene != "") %>%
    dplyr::select(term, gene)
})

key_dict <- c(
  "Astro" = "Astrocytes",
  "Endo" = "Endothelial Cells",
  "Microglia" = "Microglial Cells",
  "Neuro_Exc" = "Excitatory Neurons",
  "Neuro_Inb" = "Inhibitory Neurons",
  "Oligo" = "Oligodendrocytes",
  "OPC" = "Oligodendrocyte Progenitor Cells"
)

TERM2NAME <- TERM2GENE %>%
  dplyr::mutate(description = dplyr::recode(term, !!!key_dict)) %>%
  dplyr::select(term, description)
unique(TERM2NAME)
annoData <- build_Anno(TERM2GENE, TERM2NAME)
test1 <- get_geneSetList(annoData)
str(head(test1))
get_termDescription(term = "Astro", annoData)
saveRDS(annoData, file = "inst/extdata/geneSets/CellTypes_Martins2021.rds", compress = "xz")
```




# MeSH
First we define a function to prepare the MeSH data
```{r, include = TRUE, eval=FALSE}
get_MeSH_data <- function(MeSH.hsa.db, MeSH.db, database, category) {
  category <- toupper(category)
  categories <- c(
    "A", "B", "C", "D",
    "E", "F", "G", "H",
    "I", "J", "K", "L",
    "M", "N", "V", "Z"
  )

  if (!all(category %in% categories)) {
    stop("please check your 'category' parameter...")
  }

  mesh <- AnnotationDbi::select(MeSH.hsa.db, keys = database, columns = c("GENEID", "MESHID", "MESHCATEGORY"), keytype = "SOURCEDB")
  mesh <- mesh[mesh[, 3] %in% category, ]
  mesh2gene <- mesh[, c(2, 1)]
  mesh2name <- AnnotationDbi::select(MeSH.db, keys = unique(mesh2gene[, 1]), columns = c("MESHID", "MESHTERM"), keytype = "MESHID")
  build_Anno(mesh2gene, mesh2name)
}
```

```{r, include = TRUE, eval=FALSE}
BiocManager::install("AnnotationHub")
BiocManager::install("MeSHDbi")

library(AnnotationHub)
library(MeSHDbi)

# prepare MeSH.hsa.db (it is a gene-MeSH db)
ah <- AnnotationHub(cache = "C:/Users/Zhipeng/AppData/Local/Temp/Rtmpe24TtF/BiocFileCache", localHub = TRUE) # localHub=TRUE if you have downloaded the data
hsa <- query(ah, c("MeSHDb", "Homo sapiens"))
file_hsa <- hsa[[1]]
MeSH.hsa.db <- MeSHDbi::MeSHDb(file_hsa)

# prepare MeSH.db (it is a name db)
ah <- AnnotationHub(localHub = TRUE)
# cache="C:/Users/Zhipeng/AppData/Local/R/cache/R/AnnotationHub/205c177937a_98389"
dbfile1 <- query(ah, c("MeSHDb", "MeSH.db"))[[1]]
MeSH.db <- MeSHDbi::MeSHDb(dbfile1)

annoData <- get_MeSH_data(MeSH.hsa.db = MeSH.hsa.db, MeSH.db = MeSH.db, database = "gene2pubmed", category = "F")
test1 <- get_geneSetList(annoData)
str(head(test1))
get_termDescription(term = "D003071", annoData)

annoData <- symbol_annoData(annoData)
test2 <- get_geneSetList(annoData)
str(head(test2))
length(test2)
saveRDS(annoData, file = "inst/extdata/geneSets/test.rds", compress = "xz")
```
After figure out for one category, we can make it a loop for gene2pubmed
```{r, include = TRUE, eval=FALSE}
for (cat_i in c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L", "M", "N", "V", "Z")) {
  message("Preparing category: ", cat_i)
  # Fetching the MeSH data for the specified category
  annoData <- get_MeSH_data(MeSH.hsa.db = MeSH.hsa.db, MeSH.db = MeSH.db, database = "gene2pubmed", category = cat_i)
  geneSetList <- get_geneSetList(annoData)
  if (length(geneSetList) == 0) {
    message("No gene set found for category: ", cat_i)
  } else {
    message("Converting gene IDs to gene symbols for category: ", cat_i)
    annoData <- symbol_annoData(annoData)
    message("Saving annoData for category: ", cat_i)
    saveRDS(annoData, file = paste0("inst/extdata/geneSets/MeSH_", cat_i, ".rds"), compress = "xz")
  }
}
```


# render me
```{r include=F, eval=FALSE}
rmd_path <- "E:/xhmhc/BrainEnrich/vignettes"
rmarkdown::render(
  input = file.path(rmd_path, "prepare_annoDat.Rmd"),
  output_format = "html_document",
  output_file = file.path(rmd_path, "prepare_annoDat.html")
)
```
